### 2024-02-14 globals/metadata.tex
update ICALP

\newcommand{\makeabstract}{
    \begin{abstract}
        This paper answers two questions about 
        commutative unary output polyregular functions, called
        $\Nat$-polyregular functions, that forms a proper subset
        of $\Nat$-rational series.
        We show that given a $\Rel$-rational series,
        it is decidable whether it is computable via a commutative
        $\Nat$-polyregular function, and provide a counter-example
        to the theorem of Karhumäki that studied the same question
        in the case of polynomials.

        We also prove that it is decidable whether a commutative $\Nat$-polyregular
        function is star-free, by proving the stronger statement
        that star-free $\Rel$-polyregular functions that are
        $\Nat$-polyregular are computable using a star-free $\Nat$-polyregular function.
        This answers positively to previous conjectures on $\Nat$-polyregular functions,
        and the study of the commutative case seems to be useful
        towards a proof in the non-commutative case.
    \end{abstract}
}

%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\EventEditors{John Q. Open and Joan R. Access}
\EventNoEds{2}
\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
\EventShortTitle{ICALP 2023}
\EventAcronym{ICALP}
\EventYear{2023}
\EventDate{December 24--27, 2016}
\EventLocation{Little Whinging, United Kingdom}
\EventLogo{}
\SeriesVolume{42}
\ArticleNo{23}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

\newcommand{\makeabstract}{
    \begin{abstract}
This paper addresses two questions regarding $\Nat$-polyregular functions, that
forms a proper subset of $\Nat$-rational series. We show that given a $\Rel$-rational
series, it is decidable whether it is computable via a commutative
$\Nat$-polyregular function, and provide a counter-example to the theorem of
Karhumäki that studied the same question in the case of polynomials. We also
prove that it is decidable whether a commutative $\Nat$-polyregular function is
star-free, by proving the stronger statement that star-free $\Rel$-polyregular
functions that are $\Nat$-polyregular are in fact computable using a star-free
$\Nat$-polyregular function. Building towards answering the same questions in the
non-commutative case, we present a canonical model of computation of
$\Nat$-polyregular functions by generalizing the notion of residual transducers
previously introduced in $\Rel$-polyregular functions.
    \end{abstract}
}

%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\EventEditors{John Q. Open and Joan R. Access}
\EventNoEds{2}
\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
\EventShortTitle{ICALP 2023}
\EventAcronym{ICALP}
\EventYear{2023}
\EventDate{December 24--27, 2016}
\EventLocation{Little Whinging, United Kingdom}
\EventLogo{}
\SeriesVolume{42}
\ArticleNo{285}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2024-02-14 parts/appendix.tex
update ICALP

\begin{proof}[Proof of \cref{n-rat-correct:lem}]
    Let $f$ be the \kl{commutative}
    \kl{$\Nat$-rational series}
    with domain $\Sigma \defined \set{a_1, \dots, a_p}$
    that \kl{represents} $P$. Because $f$ has \kl{polynomial growth},
    it is in fact a \kl{$\Nat$-polyregular function}
    (\cref{polynomial-rational-polyreg:fact}).
    Using \cref{n-poly-combinatorics:lem},
    there exists a number $\omega \in \Nat$
    and polynomial $Q$ with \kl{non-negative}
    \kl{maximal monomials} such that
    for all $X_1, \dots, X_p \geq \omega$:
    \begin{equation*}
        f\left(
            \prod_{i = 1}^p a_i^{\omega X_i}
        \right)
        = Q(X_1, \dots, X_p)
        = P(\omega X_1, \dots, \omega X_p) 
        \quad .
    \end{equation*}
    We conclude that $P(\omega X_1, \dots, \omega X_p) = Q(X_1, \dots, X_p)$
    as polynomials,
    and in particular that
    the \kl{maximal monomials} of 
    $P$ are \kl{non-negative}.

@@

    Let $t$ be a first order type with at most $k$ free variables and
    quantifier rank at most $q$. The map $g_t \colon X \mapsto
    \vcount{t}(uw^X)$ is a \kl{commutative} \kl{star-free $\Nat$-polyregular
    function}. As a consequence, there exists $N_0$ and a polynomial $P \in
    \CorrectPoly$ such that for all $X \geq N_0$, $g_t(X) = P(X)$. Now, because
    $P \in \CorrectPoly$, there exists a $K \in \Nat$ such that
    $\translate{K}(\Diff{1}{P}) \in \CorrectPoly$, and in particular $g_t(X +
    N_0 + K+1) - g_t(X + N_0 + K) \in \Nat$, for all $X \geq 0$.

@@

    We can therefore define a function $\mathsf{decode} \colon \Sigma^* \to
    \sum_{q \in Q} \Nat^{d_q}$ that maps to a word $w$ to \emph{some} $q \in Q$
    and parameters $(\lambda_1, \dots, \lambda_{d_q}) \in \Nat^{d_q}$ such that
    $\mathsf{s}(w) = v_0 + \sum_{i = 1}^{d_q} \lambda_i v_{i,q}$. By taming the
    semilinear set prior to this step, we can ensure that the function is in
    fact uniquely defined, this is done by computing an equivalent semilinear
    set that is \emph{unambiguous} and \emph{proper} \cite{chistikov16}.
    Similarly, let us define $\mathsf{encode} \colon \sum_{q \in Q} \Nat^{d_q}
    \to \Sigma^*$ that maps an element $q \in Q$ with parameters $(\lambda_1,
    \dots, \lambda_{d_q}) \in \Nat^{d_q}$ to \emph{some} word $w \in \Sigma^*$
    such that $\mathsf{s}(w) = v_0 + \sum_{i = 1}^{d_q} \lambda_i v_{i,q}$.
    Notice that $\mathsf{decode} \circ \mathsf{encode} = \mathsf{id}$, by the
    unicity of the representation in $\sum_{q \in Q} \Nat^{d_q}$.
    We can define the map $\mathsf{repr} \colon \sum_{q \in Q} \Nat^{d_q} \to \Nat^{M^3}$
    as follows
    \begin{equation*}
        \mathsf{repr}(q, \lambda_1, \dots, \lambda_{d_q}) \defined
        v_0 + \sum_{i = 1}^{d_q} \lambda_i v_{i,q}
        \quad .
    \end{equation*}
    Finally, let us define for $q \in Q$ the map $g_q \colon \Nat^{d_q} \to \Rel$ as the polynomial
    expression:
    \begin{equation*}
        g_q(\lambda_1, \dots, \lambda_{d_q}) \defined \mathsf{repr}(q, \lambda_1, \dots, \lambda_{d_q}) \quad .
    \end{equation*}

@@

    The only missing part of this proof is the fact that $\mathsf{encode}$,
    $\mathsf{decode}$ can be computed as \kl{polyregular functions}. Notice
    that \emph{mutatis mutandis}, the same proof scheme applies to $f \in
    \ZSF[k]$, by assuming that the monoid is aperiodic, since the functions
    constructed (apart from $\mathsf{encode}$ and $\mathsf{decode}$) will be
    \kl{star-free}.
\end{proof}

>>>>

\begin{proof}[Proof of \cref{n-rat-correct:lem}]
    Let $f$ be the \kl{commutative}
    \kl{$\Nat$-rational series}
    with domain defined as $\Sigma \defined \set{a_1, \dots, a_p}$
    that \kl{represents} $P$. Because $f$ has \kl{polynomial growth},
    it is in fact a \kl{$\Nat$-polyregular function}
    (\cref{polynomial-rational-polyreg:fact}).
    Using \cref{n-poly-combinatorics:lem},
    there exists a number $\omega \in \Nat$
    and polynomial $Q$ with \kl{non-negative}
    \kl{maximal monomials} such that
    for all $X_1, \dots, X_p \geq \omega$:
    \begin{equation*}
        f\left(
            \prod_{i = 1}^p a_i^{\omega X_i}
        \right)
        = Q(X_1, \dots, X_p)
        = P(\omega X_1, \dots, \omega X_p) 
        \quad .
    \end{equation*}
    We conclude that $P(\omega X_1, \dots, \omega X_p) = Q(X_1, \dots, X_p)$
    as polynomials,
    and in particular that
    the \kl{maximal monomials} of 
    $P$ are \kl{non-negative}.

@@

    Let $t$ be a first order type with at most $k$ free variables and
    quantifier rank at most $q$. The map $g_t \colon X \mapsto
    \vcount{t}(uw^X)$ is a \kl{commutative} \kl{star-free $\Nat$-polyregular
    function}. As a consequence, there exists $N_0$ and a polynomial $P \in
    \CorrectPoly$ such that for all $X \geq N_0$, $g_t(X) = P(X)$. Now, because
    $P \in \CorrectPoly$, there exists a $K \in \Nat$ such that
    $\translate{K}(\Diff{1}{P}) \in \Nat[X]$
    (\cref{derivation-translation:lem}), and in particular $g_t(X + N_0 + K+1)
    - g_t(X + N_0 + K) \in \Nat$, for all $X \geq 0$.

@@

    We can therefore define a function $\mathsf{decode} \colon \Sigma^* \to
    \sum_{q \in Q} \Nat^{d_q}$ that maps to a word $w$ to \emph{some} $q \in Q$
    and parameters $(\lambda_1, \dots, \lambda_{d_q}) \in \Nat^{d_q}$ such that
    $\mathsf{s}(w) = v_0 + \sum_{i = 1}^{d_q} \lambda_i v_{i,q}$. By taming the
    semilinear set prior to this step, we can ensure that the function is in
    fact uniquely defined, this is done by computing an equivalent semilinear
    set that is \emph{unambiguous} and \emph{proper} \cite{chistikov16}.
    Similarly, let us define $\mathsf{encode} \colon \sum_{q \in Q} \Nat^{d_q}
    \to \Sigma^*$ that maps an element $q \in Q$ with parameters $(\lambda_1,
    \dots, \lambda_{d_q}) \in \Nat^{d_q}$ to \emph{some} word $w \in \Sigma^*$
    such that $\mathsf{s}(w) = v_0 + \sum_{i = 1}^{d_q} \lambda_i v_{i,q}$.
    Notice that $\mathsf{decode} \circ \mathsf{encode} = \mathsf{id}$, by the
    unicity of the representation in $\sum_{q \in Q} \Nat^{d_q}$.
    We can define the map $\mathsf{repr} \colon \sum_{q \in Q} \Nat^{d_q} \to \Nat^{M^3}$
    as follows:
    \begin{equation*}
        \mathsf{repr}(q, \lambda_1, \dots, \lambda_{d_q}) \defined
        v_0 + \sum_{i = 1}^{d_q} \lambda_i v_{i,q}
        \quad .
    \end{equation*}
    Finally, let us define for $q \in Q$ the map $g_q \colon \Nat^{d_q} \to \Rel$ as the polynomial
    expression:
    \begin{equation*}
        g_q(\lambda_1, \dots, \lambda_{d_q}) \defined \mathsf{repr}(q, \lambda_1, \dots, \lambda_{d_q}) \quad .
    \end{equation*}

@@

    The only missing part of this proof is the fact that the maps $\mathsf{encode}$ and
    $\mathsf{decode}$ can be computed as \kl{polyregular functions}. Notice
    that \emph{mutatis mutandis}, the same proof scheme applies to $f \in
    \ZSF[k]$, by assuming that the monoid is aperiodic, since the functions
    constructed (apart from $\mathsf{encode}$ and $\mathsf{decode}$) will be
    \kl{star-free}.
\end{proof}

### 2024-02-14 parts/correction.tex
update ICALP

In the actual statement of \cref{corrected-version:thm}, the representation of
polynomials in $\CorrectPoly$ is strengthened from \kl{$\Nat$-polyregular
functions} to \kl{star-free $\Nat$-polyregular functions}. This is not
surprising, given the conjecture that \kl{ultimately polynomial}
\kl{$\Nat$-polyregular functions} (re-introduced in \cref{star-free:sec}) were
conjectured to be star free (\cref{zsf-nsf:conjecture}), and that polynomials
are always \kl{ultimately polynomial}.

\AP When multiple indeterminate are involved, the notion of degree must be
adapted, and to ensure that the decision procedure is effective, one must
obtain explicit bounds on the translations that are needed. To that end, let us
now introduce the ordering over which the induction of
\cref{lem:correct-to-n-rat} is built. Recall that $\MaximalMonomials(P)$ is the
set of \kl{maximal monomials} of $P$, hence belongs to $\Pfin(\Monomials)$. We
order \kl{monomials} with the \kl{divisibility ordering}, making $\Monomials$ a
\kl{well-quasi-ordering} that is isomorphic to $\Nat^k$ with the product
ordering \cite[Dickson’s Lemma]{SCSC12}. We endow $\Pfin(\Monomials)$ with the
\intro{Hoare ordering}, that is, $S_1 \hoareleq S_2$ whenever for all
\kl{monomials} $M_1 \in S_1$, there exists a monomial $M_2 \in S_2$, such that
$M_1$ \kl{divides} $M_2$. The set $(\Pfin(\Monomials), \hoareleq)$ remains a
\kl{well-quasi-ordering} \cite[Hoare quasi-ordering]{SCSC12}.

>>>>

In the actual statement of \cref{corrected-version:thm}, the representation of
polynomials in $\CorrectPoly$ is strengthened from \kl{$\Nat$-polyregular
functions} to \kl{star-free $\Nat$-polyregular functions}. This is not
surprising, given the conjecture that \kl{ultimately polynomial}
\kl{$\Nat$-polyregular functions} (the formal definition will be given in
\cref{star-free:sec}) were conjectured to be star free
(\cref{zsf-nsf:conjecture}), and that polynomials are always \kl{ultimately
polynomial}.

\AP When multiple indeterminate are involved, the notion of degree must be
adapted, and to ensure that the decision procedure is effective, one must
obtain explicit bounds on the \kl{translations} that are needed. To that end,
let us now introduce the ordering over which the induction of
\cref{lem:correct-to-n-rat} is built. Recall that $\MaximalMonomials(P)$ is the
set of \kl{maximal monomials} of $P$, hence belongs to $\Pfin(\Monomials)$. We
order \kl{monomials} with the \kl{divisibility ordering}, making $\Monomials$ a
\kl{well-quasi-ordering} that is isomorphic to $\Nat^k$ with the product
ordering \cite[Dickson’s Lemma]{SCSC12}. We endow $\Pfin(\Monomials)$ with the
\intro{Hoare ordering}, that is, $S_1 \hoareleq S_2$ whenever for all
\kl{monomials} $M_1 \in S_1$, there exists a monomial $M_2 \in S_2$, such that
$M_1$ \kl{divides} $M_2$. The set $(\Pfin(\Monomials), \hoareleq)$ remains a
\kl{well-quasi-ordering}, and is in particular well-founded \cite[Hoare
quasi-ordering]{SCSC12}.

### 2024-02-14 parts/intro.tex
update ICALP

Traditional results from automata theory state the equivalence of languages
defined in terms of deterministic finite automata and regular expressions
\cite{kleene1956representation}, Monadic Second Order Logic ($\intro*\MSO$)
\cite{buchi1960weak}, or finite monoids \cite{schutzenberger1961definition}.
Among regular languages, there exists a robust subclass sharing similar
characterizations, namely the \emph{star-free} regular languages. These are
obtained via counter-free (minimal) automata \cite{mcnaughton1971counter},
represented by star-free regular expressions \cite{schutzenberger1965finite},
definable in First Order Logic ($\intro*\FO$) \cite{perrin1986first}, and recognized
by aperiodic monoids \cite{schutzenberger1965finite}. It is decidable whether a
regular language is aperiodic, thanks to the characterizations in terms of
minimal automaton and/or syntactic monoid.

There have been recurrent attempts to generalize the notions of regularity from
languages to functions, leading to classes of various expressiveness
(sequential, rational, regular, and \kl{polyregular functions}). In this paper,
we will focus on \intro{polyregular functions} ($\intro*\Poly$), that shares with regular
languages its characterizations in terms of logic (via $\MSO$ interpretations),
automata (via pebble automata), and adds new bridges with programming languages
via its relationships with lambda-terms and so-called for-transducers
\cite{bojanczyk2019string}. As for regular languages, the class
$\intro*\SF$ of \intro{star-free polyregular function} can equivalently be introduced
in terms of first-order definability, counter-free pebble transducers, and
suitable restrictions of for-tranducers.

Classical problems, including the decidability of aperiodicity, become
difficult in the case of \kl{polyregular functions}, due to the lack of
canonical objects akin to the syntactic monoid or the minimal automaton. For
sequential functions (that admit a canonical object), aperiodicity is decidable
\cite{choffrut03}, and more recently, aperiodicity has been proven decidable
for rational functions \cite{filiot2016aperiodicity,filiot2018canonical}. The
problem remains open for regular functions, although it has been proven
decidable when changing the semantics of function equality to account for the
\emph{origins} in the output \cite{bojanczyk14}. To construct subclasses of
\kl{polyregular function}, an interesting parameter is the \intro{growth rate}
of a function $f$, which is given by a number $k \in \Nat$, which can be used
to create a strict hierarchy inside $\Poly$. All functions in $\Poly$ have
\intro{polynomial growth}, and we can write $\Poly[k]$ for the \kl{polyregular
functions} of \kl{growth rate} $k$. It has been shown that the parameter $k$
controls the maximal complexity of the $\MSO$ interpretation needed to
represent the function \cite{bojanczyk2022transducers}.

Another way to approach the problem of deciding if a \kl{polyregular function}
is \kl{star-free}, is to not restrict the computational power (via the
\kl{growth rate}), nor change the semantics (by adding origins), but consider
restrictions of the output alphabet. In particular, there have been successful
results proven on \emph{unary polyregular functions}, also known as
\kl{$\Nat$-polyregular functions} ($\intro*\NPoly$)
\cite{doueneau2021pebble,doueneau2022hiding}. For instance, the \kl{growth
rate} $k$ of a function $f \in \NPoly$ provides the minimal number of pebbles
needed to compute $f$ \cite{doueneau2021pebble}, which does not hold on $\Poly$
\cite{bojanczyk2022transducers}. Let us remark that the restriction to
\emph{unary output} in the case of \kl{polyregular functions} can be lifted to
outputs in the free commuative semigroup generated by a finite alphabet
$\Gamma$. In this sense, the shift to \emph{unary output} is really about
enforcing commutativity on the output.

Recently, aperiodicity has been shown decidable over a restricted class of
\kl{polyregular functions} where the ouptut belongs to $\Rel$ instead of
$\Nat$, that are called \kl{$\Rel$-polyregular functions} ($\intro*\ZPoly$)
\cite{LOPEZ23b}. These functions generalize \kl{$\Nat$-polyregular functions}
by allowing negative outputs, and therefore enabling a limited form of
backtracking. As for the unary outputs, these results generalize to
\kl{polyregular functions} having outputs in the free commutative \emph{group}
generated by a finite alphabet $\Gamma$. The decidability of aperiodicity for
\kl{$\Rel$-polyregular functions} relies on the construction of a canonical
object called the \emph{residual transducer}, the latter being essentially
based on differences between functions, crucially leveraging negative outputs
(i.e., the presence of inverse elements). Although the proof method does not carry
from $\Rel$-output functions to $\Nat$-output functions, it was conjectured
that the semantic characterization of \kl{star-free $\Rel$-polyregular
functions} ($\intro*\ZSF$) would continue to characterize \kl{star-free
$\Nat$-polyregular functions} ($\intro*\NSF$).

@@

It is well known that $\NPoly$ (resp. $\ZPoly$) is proper subclass respectively
of \intro{$\Nat$-rational series} ($\intro*\NRat$) (resp. of \intro{$\Rel$-rational
series} $\intro*\ZRat$), that have been characterized in terms of \kl{growth rate},
ambiguity, and a notion of \emph{quantitative} $\MSO$
\cite{schutzenberger1962,kreutzer2013}, and therefore can benefit from generic
theorems obtained on these \cite{berstel2011noncommutative}. However, for these
larger classes of functions, several competing notions of aperiodicity exist
\cite{droste2019aperiodic,reutenauer_series_1980,LOPEZ23b}, and the
decidability of the ones that extends $\ZSF$ and $\NSF$ remains open.
Furthermore, the decidability of $\NRat$ inside $\ZRat$ is also a longstanding
open problem, that was to the knowledge of the authors, only solved by
\textcite{KARH77} in the case of \kl{$\Rel$-rational series} that compute
\emph{polynomials}, with an incorrect proof. Furthermore, decidability of
$\NPoly$ inside $\ZPoly$ (resp. $\NSF$ inside $\ZSF$) remains an open problem
\cite[Open question 5.55]{gaetanphd}.

We refer the reader to \cref{previously-known-inclusions:fig} to get an
overview of the different classes of functions appearing in the paper. Arrows
denote strict inclusions, and effectiveness (both in terms of decidability and
of effective representation) is represented by thick double arrows. Inclusions
that are suspected to be effective are represented using a dashed arrow
together with a question mark.

@@

>>>>

\AP
Traditional results from automata theory state the equivalence of languages
defined in terms of deterministic finite automata and regular expressions
\cite{kleene1956representation}, Monadic Second Order Logic ($\intro*\MSO$)
\cite{buchi1960weak}, or finite monoids \cite{schutzenberger1961definition}.
Among regular languages, there exists a robust subclass sharing similar
characterizations, namely the \emph{star-free} regular languages. These are
obtained via counter-free (minimal) automata \cite{mcnaughton1971counter},
represented by star-free regular expressions \cite{schutzenberger1965finite},
definable in First Order Logic ($\intro*\FO$) \cite{perrin1986first}, and recognized
by aperiodic monoids \cite{schutzenberger1965finite}. It is decidable whether a
regular language is aperiodic, thanks to the characterizations in terms of
minimal automaton and/or syntactic monoid.

\AP There have been recurrent attempts to generalize the notions of regularity
from languages to functions, leading to classes of various expressiveness (for
instance, sequential, rational, regular, and \kl{polyregular functions})
\cite{EM65,Aho1969,eilenberg1974automata,EM02}. In this paper, we will focus on
\intro{polyregular functions} ($\intro*\Poly$), that shares with regular
languages its characterizations in terms of logic (via $\MSO$ interpretations),
automata (via pebble automata), and adds new bridges with programming languages
via its relationships with lambda-terms and so-called for-transducers
\cite{bojanczyk2019string}. As for regular languages, the class $\intro*\SF$ of
\intro{star-free polyregular function} can equivalently be introduced in terms
of first-order definability, counter-free pebble transducers, and suitable
restrictions of for-tranducers.

\AP
Classical problems, including the decidability of aperiodicity, become
difficult in the case of \kl{polyregular functions}, due to the lack of
canonical objects akin to the syntactic monoid or the minimal automaton. For
sequential functions (that admit a canonical object), aperiodicity is decidable
\cite{choffrut03}, and more recently, aperiodicity has been proven decidable
for rational functions \cite{filiot2016aperiodicity,filiot2018canonical}. The
problem remains open for regular functions, although it has been proven
decidable when changing the semantics of function equality to account for the
\emph{origins} in the output \cite{bojanczyk14}. To construct subclasses of
\kl{polyregular function}, an interesting parameter is the \intro{growth rate}
of a function $f$, which is given by a number $k \in \Nat$, which can be used
to create a strict hierarchy inside $\Poly$. All functions in $\Poly$ have
\intro{polynomial growth}, and we can write $\Poly[k]$ for the \kl{polyregular
functions} of \kl{growth rate} $k$. It has been shown that the parameter $k$
controls the maximal complexity of the $\MSO$ interpretation needed to
represent the function \cite{bojanczyk2022transducers}.

\AP
Another way to approach the problem of deciding if a \kl{polyregular function}
is \kl{star-free}, is to not restrict the computational power (via the
\kl{growth rate}), nor change the semantics (by adding origins), but consider
restrictions of the output alphabet. In particular, there have been successful
results proven on \emph{unary polyregular functions}, also known as
\kl{$\Nat$-polyregular functions} ($\intro*\NPoly$)
\cite{doueneau2021pebble,doueneau2022hiding}. For instance, the \kl{growth
rate} $k$ of a function $f \in \NPoly$ provides the minimal number of pebbles
needed to compute $f$ \cite{doueneau2021pebble}, which does not hold on $\Poly$
\cite{bojanczyk2022transducers}. Let us remark that the restriction to
\emph{unary output} in the case of \kl{polyregular functions} can be lifted to
outputs in the free commuative semigroup generated by a finite alphabet
$\Gamma$. In this sense, the shift to \emph{unary output} is really about
enforcing commutativity on the output.

\AP
Recently, aperiodicity has been shown decidable over a restricted class of
\kl{polyregular functions} where the ouptut belongs to $\Rel$ instead of
$\Nat$, that are called \kl{$\Rel$-polyregular functions} ($\intro*\ZPoly$)
\cite{LOPEZ23b}. These functions generalize \kl{$\Nat$-polyregular functions}
by allowing negative outputs, and therefore enabling a limited form of
backtracking. As for the unary outputs, these results generalize to
\kl{polyregular functions} having outputs in the free commutative \emph{group}
generated by a finite alphabet $\Gamma$. The decidability of aperiodicity for
\kl{$\Rel$-polyregular functions} relies on the construction of a canonical
object called the \emph{residual transducer}, the latter being essentially
based on differences between functions, crucially leveraging negative outputs
(i.e., the presence of inverse elements). Although the proof method does not carry
from $\Rel$-output functions to $\Nat$-output functions, it was conjectured
that the semantic characterization of \kl{star-free $\Rel$-polyregular
functions} ($\intro*\ZSF$) would continue to characterize \kl{star-free
$\Nat$-polyregular functions} ($\intro*\NSF$).

@@

\AP
It is well known that $\NPoly$ (resp. $\ZPoly$) is proper subclass respectively
of \intro{$\Nat$-rational series} ($\intro*\NRat$) (resp. of \intro{$\Rel$-rational
series} $\intro*\ZRat$), that have been characterized in terms of \kl{growth rate},
ambiguity, and a notion of \emph{quantitative} $\MSO$
\cite{schutzenberger1962,kreutzer2013}, and therefore can benefit from generic
theorems obtained on these \cite{berstel2011noncommutative}. However, for these
larger classes of functions, several competing notions of aperiodicity exist
\cite{droste2019aperiodic,reutenauer_series_1980,LOPEZ23b}, and the
decidability of the ones that extends $\ZSF$ and $\NSF$ remains open.
Furthermore, the decidability of $\NRat$ inside $\ZRat$ is also a longstanding
open problem, that was to the knowledge of the authors, only solved by
\textcite{KARH77} in the case of \kl{$\Rel$-rational series} that compute
\emph{polynomials}, with an incorrect proof. Furthermore, decidability of
$\NPoly$ inside $\ZPoly$ (resp. $\NSF$ inside $\ZSF$) remains an open problem
\cite[Open question 5.55]{gaetanphd}.

\AP
We refer the reader to \cref{previously-known-inclusions:fig} to get an
overview of the different classes of functions appearing in the paper. Arrows
denote strict inclusions, and effectiveness (both in terms of decidability and
of effective representation) is represented by thick double arrows. Inclusions
that are suspected to be effective are represented using a dashed arrow
together with a question mark.

@@

The approach taken in this paper avoids leveraging tools from the theory of
factorizations forests in finite monoids \cite{simon90}, which has been the
traditional method used to understand (commutative output) \kl{polyregular
functions} \cite{doueneau2021pebble,doueneau2022hiding,LOPEZ23b}, and instead
focus on intrinsic description of the functions themselves (via canonical
computational models such as the \kl{residual transducer} of
\cref{residual-transducer:def}, or explicit encoding/decoding into polynomials
via \cref{decompose-polynomial:lem}).

### 2024-02-14 parts/prelim.tex
update ICALP

\AP In order to build functions in $\NPoly$, the following constructions will
be useful. For every language $L \subseteq \Sigma^*$ we define $\ind{L} \colon
\Sigma^* \to \set{0,1}$ as the indicator function of $L$. Given an $\MSO$
formula with first-order free variables $\varphi(\vec{x})$, we define
$\vcount{\varphi(\vec{x})}$ as the function from $\Sigma^*$ to $\Nat$ that maps
a word $w$ to the number of assignments $\nu \colon \vec{x} \to w$, such that
$w, \nu \models \varphi(\vec{x})$. If $X$ is a set of functions from $\Sigma^*$
to $\Rel$, and a subset $\mathbb{S}$ of $\Rel$, we define
$\Span{\mathbb{S}}(X)$ as the smallest subset of functions from $\Sigma^* \to
\Rel$ containing $X$ and closed under (finite) sums and multiplication with $s
\in \mathbb{S}$.

@@

\AP All functions in this paper have output in $\Nat$ or $\Rel$, hence, we
avoid the definition of \kl{polyregular functions} in full generality. The
following \cref{nat-rel-poly:def} is one of the equivalent definitions of
\cite{LOPEZ23b}, and is similar in shape to the \emph{finite counting automata}
introduced by \textcite{schutzenberger1962}. Let us recall that a monoid $M$ is
\intro(monoid){aperiodic} whenever there exists $n \in \Nat$ such that for all
$x \in M$, $x^{n+1} = x^n$.

@@

\AP Let us briefly recall in the following lemma alternative characterizations
of $\NPoly$ (resp. $\ZPoly$), some of which will be helpful in the upcoming
analysis of their \kl{commutative} counterpart, and others that provide
a nicer syntax to construct examples. These characterizations were previously known, and
can be found in \cite{gaetanphd,LOPEZ23b}.

@@

\AP Because this paper focuses on functions that are commutative both in the
input and the output, let us introduce precisely what it means and how it
relates to other functions from $\Nat^k \to \Nat$. For that, let $\Sigma$ be a
finite alphabet. We define $\commute \colon \Sigma^* \to \Nat^\Sigma$ for the
map that counts occurrences of every letter in the input word (that is,
computes the Parikh vector) namely: $ \commute[w] \defined (a \mapsto
\card[a]{w})$. Let $X$ be a set, a function $f \colon \Sigma^* \to X$ is
\intro{commutative} whenever for all $u,v \in \Sigma^*$, $f(uv) = f(vu)$.
Equivalently, it is \reintro{commutative} whenever there exists a map $g \colon
\Nat^\Sigma \to X$ such that $g \circ \commute = f$.

@@

\AP A function $g \colon \Nat^n \to \Rel$ is \intro{represented} by a function
$f \colon \Sigma^* \to \Rel$ if there exists a map $\eta \colon \set{1, \dots,
n} \to \Sigma$ such that $g \circ (v \mapsto \seqof{v_{\eta(k)}}{1 \leq k \leq
n}) \circ \commute = f$. A key family of functions will be \kl{represented} by
\kl{$\Rel$-polyregular functions}, namely polynomials with multiple indeterminates.

@@

All \kl{$\Nat$-rational polynomials} are \kl{non-negative}, but the converse does
not hold,
as the following example illustrates. Note that the proof scheme of
this example will be at the core of \cref{thm:counter-example}, and leverages
a key property of $\NRat$ recalled hereafter.

>>>>

\AP In order to build functions in $\NPoly$, the following constructions will
be useful. For every language $L \subseteq \Sigma^*$ we define $\ind{L} \colon
\Sigma^* \to \set{0,1}$ as the indicator function of $L$. Given an $\MSO$
(resp. $\FO$) formula with first-order free variables $\varphi(\vec{x})$, we
define $\vcount{\varphi(\vec{x})}$ as the function from $\Sigma^*$ to $\Nat$
that maps a word $w$ to the number of assignments $\nu \colon \vec{x} \to w$,
such that $w, \nu \models \varphi(\vec{x})$. If $X$ is a set of functions from
$\Sigma^*$ to $\Rel$, and a subset $\mathbb{S}$ of $\Rel$, we define
$\intro*\Span{\mathbb{S}}(X)$ as the smallest subset of functions from
$\Sigma^* \to \Rel$ containing $X$ and closed under (finite) sums and
multiplication with $s \in \mathbb{S}$.

@@

\AP Because the functions of interest in this paper have output in $\Nat$ or
$\Rel$, we will only provide the definiton of \kl{polyregular functions} for
these two output semgroups, and we refer the reader to
\cite{bojanczyk2019string} for the general definition of \kl{polyregular
functions}. The following \cref{nat-rel-poly:def} is one of the equivalent
definitions of \cite{LOPEZ23b}, and is similar in shape to the \emph{finite
counting automata} introduced by \textcite{schutzenberger1962}. Let us recall
that a monoid $M$ is \intro(monoid){aperiodic} whenever there exists $n \in
\Nat$ such that for all $x \in M$, $x^{n+1} = x^n$.

@@

\AP Let us briefly recall in the following lemma alternative characterizations
of $\NPoly$ (resp. $\ZPoly$), some of which will be helpful in the upcoming
analysis of their \kl{commutative} counterpart, and others that provide a nicer
syntax to construct examples. These characterizations were previously known,
and can be found for instance in \cite{gaetanphd,LOPEZ23b}.

@@

\AP 
Let $\Sigma$ be a finite alphabet. 
We define $\commute \colon \Sigma^* \to \Nat^\Sigma$ for the
map that counts occurrences of every letter in the input word (that is,
computes the Parikh vector) namely: $ \commute[w] \defined (a \mapsto
\card[a]{w})$. Let $X$ be a set, a function $f \colon \Sigma^* \to X$ is
\intro{commutative} whenever for all $u,v \in \Sigma^*$, $f(uv) = f(vu)$.
Equivalently, it is \reintro{commutative} whenever there exists a map $g \colon
\Nat^\Sigma \to X$ such that $g \circ \commute = f$.

@@

\AP A function $g \colon \Nat^n \to \Rel$ is \intro{represented} by a function
$f \colon \Sigma^* \to \Rel$ if there exists a map $\eta \colon \set{1, \dots,
n} \to \Sigma$ such that $g \circ (v \mapsto \seqof{v_{\eta(k)}}{1 \leq k \leq
n}) \circ \commute = f$. This notion will be useful to formally state that a
polynomial ``is'' a \kl{polyregular function}.

@@

All \kl{$\Nat$-rational polynomials} are \kl{non-negative}, but the converse
does not hold, as the following example illustrates. Note that the proof scheme
of this example will be at the core of the counter-example to \cite[Theoerm
3.3]{KARH77} in \cref{thm:counter-example}, and leverages a key property of
$\NRat$ recalled hereafter.
